Topological Equivariant Artist Model for Visualization Library Architecture

### 2024-08-22 paper/main.tex
figures cleanup to verify they're showing what supposed to

\IEEEPARstart{V}{isualization} tools implement transforms of the form $data \rightarrow graphic$ and are expected to be structure preserving. Loosely, that means the mathematical properties of the data and how the points are connected to each other should be inferable such that  $graphic \rightarrow data$. For example, values read off a bar chart have to be equivalent to the values used to construct that chart. Therefore a visualization tool is structure preserving when it preserves the bidirectional mapping $data\leftrightarrow graphic$.

@@

 In \autoref{fig:related-work:continuity:ktypes}, the topology is a continuous 3 dimensional surface encapsulating time and position, while the fields are date, location, station, pressure, and temperature. Our paper provides a language to express how the scatter plot, line plot, and heatmap are preserving different structural properties of the dataset, thereby generalizing Bertin's codification of how the topology of observations must match the class of representation (i.e. point , line, area) and how the variation possible in a field match the variation in the corresponding visual variable \cite{bertinSemiologyGraphicsDiagrams2011} and Mackinlay's work on how structure preservation includes preserving the mathimatical relations between elements of the same and different fields \cite{mackinlayAutomatingDesignGraphical1986}.

@@

Computer science data types roughly describe structure in an equivalent manner;
objects are elements of the set and class operators and methods are akin to
actions, as described in \autoref{sec:atct:fb:fiber}. Given this equivalancy between code and math, we can develop software functions to test whether structure is preserved based on mathematical tests of structure preservation. For example, \textit{equivariance} allows us to test stucture preservation under transformations to data or visualization and has been proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014} and \textit{homomorphism} provides a way of asserting whether the relations between data elements was preserved as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}.

\subsubsection{equivariance}
Specifically, Steven's conceptualizes the structure on values as \textcolor{action}{actions} on groups \footnote{A \textit{group} is a set with an associative binary operator. This operation must have an identity element and be closed, associative,
and invertable}. A function that preserves structure when the input or output is changed by a group action is called \textit{equivariant}. Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$

@@

\subsubsection{homomorphism}

For example, although the Steven's scales were conceptualized as having group structure, the ordinal scale have a monoidal structure because partial orders ($\geq, \leq$) are not invertable. This means we cannot use \textit{equivariance} to test for structure preservation instead we use \textit{homomorphsim} because it imposes fewer constraints on the underlying mathematical structure of the data.

>>>>

\IEEEPARstart{V}isualization design guidelines, generally, describe how to choose visual encodings that preserve the structure of the data; to follow these guidelines the visualization tools that implement these $\data \rightarrow graphic$ transforms must be structure preserving. Loosely, preserving structure means that the athematical properties of the data and how the points are connected to each other should be inferable from the graphic such that  a $graphic \rightarrow data$ mapping can be made. For example, values read off a bar chart have to be equivalent to the values used to construct that chart. Therefore a visualization tool is structure preserving when it preserves the bidirectional mapping $data\leftrightarrow graphic$.

@@

\note{this got moved}
In \autoref{fig:related-work:continuity:ktypes}, the topology is a continuous 3 dimensional surface encapsulating time and position, while the fields are date, location, station, pressure, and temperature. Our paper provides a language to express how the scatter plot, line plot, and heatmap are preserving different structural properties of the dataset, thereby generalizing Bertin's codification of how the topology of observations must match the class of representation (i.e. point , line, area) and how the variation possible in a field match the variation in the corresponding visual variable \cite{bertinSemiologyGraphicsDiagrams2011} and Mackinlay's work on how structure preservation includes preserving the mathimatical relations between elements of the same and different fields \cite{mackinlayAutomatingDesignGraphical1986}.

@@

The elements of a set $X$ are roughly equivalent to instantiated objects of a class, and actions are roughly equivalent to methods on the objects; therefore mathematical definitions of structure preservation can be mapped directly to code. For example, \textit{equivariant} functions preserve structure under transformations to data or visualization and has been proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014} and \textit{homomorphic} maps preserve relations between data elements was preserved as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}.

\subsubsection{equivariance}
\begin{figure}
  \includegraphics*[width=1\columnwidth]{tex/equivariant.png}
\end{figure}
Specifically, Steven's conceptualizes the structure on values as \textcolor{action}{actions} on groups \footnote{A \textit{group} is a set with an associative binary operator. This operation must have an identity element and be closed, associative,
and invertable}. A function that preserves structure when the input or output is changed by a group action is called \textit{equivariant}. Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$

@@

\subsubsection{homomorphism}
\begin{figure}
  \includegraphics[width=1\columnwidth]{tex/homomorphism.png}
\end{figure}
For example, although the Steven's scales were conceptualized as having group structure, the ordinal scale have a monoidal structure because partial orders ($\geq, \leq$) are not invertable. This means we cannot use \textit{equivariance} to test for structure preservation instead we use \textit{homomorphsim} because it imposes fewer constraints on the underlying mathematical structure of the data.
